TCP Rax: Deep Reinforcement Learning for Congestion Control

### 2018-01-31 tex/paper.tex
Add all results data and finish tex

%Conference
\acmConference[WOODSTOCK'97]{ACM Woodstock conference}{July 1997}{El
  Paso, Texas USA} 
\acmYear{2018}
\copyrightyear{2018}

@@

\begin{document}
\title{TCP Rax: Deep Reinforcement Learning for TCP Congestion Control}
%\titlenote{Produces the permission block, and
%  copyright information}
%\subtitle{Extended Abs\texttt{tract}
%\subtitlenote{The full version of the author's guide is available as
%  \texttt{acmart.pdf} document}}

@@

@@

This paper proposes Reactive Adaptive eXperience based congestion control (TCP Rax), a new method of congestion control that uses online reinforcement learning to maintain an optimum congestion window with respect to a given utility function and based on current network conditions. We use a neural network based approach that can be initialized either with random weights (without prior knowledge) or with a previously trained neural network to improve stability and convergence time. The processing of rewards in congestion control depends on the arrival of acknowledgments, inducing loop delays that may lead to oscillations. As a remedy we propose \textit{Partial Action Learning}, a specific formulation of neural network based reinforcement learning that supports the processing of delayed and partial rewards.

We show that our method converges to a stable, close-to-optimum solution within the order of minutes and outperforms existing congestion control algorithms in typical networks. Thus, this paper demonstrates that neural network based reinforcement learning without any prior knowledge can feasibly be done on-line and can compete with human-designed congestion-control schemes such as New Reno and Cubic.

@@

In general, in congestion control each sender aims to maximize an objective function that usually comprises one or more of the following subgoals:
\begin{enumerate*}
\item Send as many packets as possible
\item with as little delay as possible
\item while minimizing the amount of lost packets
\end{enumerate*}. 
As, for instance, for some applications overall throughput is more important than little packet loss, each application can define its preference using a custom utility function. Besides aiming to fulfill these subgoals, it is also desirable from an overall point of view that all senders use available resources in a way that is fair to the others: When several senders share an Internet link under the same conditions, they should receive an equal share of the available resources. To achieve fairness, it is necessary that senders only use utility functions that 
\begin{enumerate*}
\item do not give an unfair advantage over others
\item and converge to a stable equilibrium
\end{enumerate*}.

In a machine learning based congestion control method, each sender uses a set of observed environment conditions to determine which action it should take to maximize his reward in the future. The environment conditions can be any metrics that the sender can obtain from the network, for example the mean round-trip time of the last received packets, the packet loss rate etc. An action is a change to the sending rate: For example, if a sender perceives an increase in the loss rate it might be advisable to lower the sending rate. To modulate the sending rate one can  either
\begin{enumerate*}
\item change the sending rate itself at certain points in time or
\item maintain a congestion window that indicates the number of packets that can be in flight at a certain point in time
\end{enumerate*}. It seems preferable to maintain a congestion window, as one cannot be certain that the operating system (OS) can maintain a certain constant sending rate for each application on one machine, as the OS has to schedule different applications and share the network interfaces between them.

While congestion control seems to be a well-suited problem for reinforcement learning, there are a couple of unique aspects and differences that distinguish it from most other reinforcement learning problems:  \begin{itemize}
\item \textbf{Delayed rewards}: In congestion control, rewards are always delayed by one round-trip time, while in most other areas where reinforcement learning is applied, if one performs an action, one can see the consequences of it action immediately. Also, due to the delay, by the time an action receives its reward, probably other actions have already been carried out in the meantime. Thus, contrary to most RL applications, actions and rewards are not synchronized. 
\item \textbf{Actions and rewards are not atomic}: If an action causes three packets to be sent, we consider these three packets being sent three \textit{partial actions}. In case these packets are transmitted correctly, the receiver will send back three acknowledgements. Each of these acknowledgements is a \textit{partial reward} that already unveils new information on the current state of the network and enables the sender to take an action based on this new information. However, only if all partial rewards have been received, the sender can assemble the overall reward of the corresponding action and can perform an update of the underlying reinforcement learning logic. 
\item \textbf{Variable number of actions and rewards}: If the congestion window is larger, more packets are sent and more acknowledgements are received, which means that the reinforcement learning algorithm gets more training data the larger the window is. This is problematic as it means that -- without proper adjustments -- the experience made at a larger congestion window will ``overwrite'' the experience made at a smaller window as with the larger window simply more training data arrives and it is thus going to be overemphasized. 
\end{itemize}

@@

While using PAL to train an optimum congestion control for a specific range of network scenarios in an off-line fashion is possible, offline learning of an optimum TCP congestion control is something that has already been done previously in an approach called \textit{Remy} \citep{winstein_tcp_2013}. The only potential advantage that PAL could provide over these previous solutions here is increased training speed. Thus, we want to show that it is  not only possible to learn congestion control by using machine learning but that it is even possible to do so on-line and without any preknowledge about the network environment.

Using a utility function that encourages high throughput and discourages packet loss (adapted from \citep{dong_pcc:_2015}) we see that our machine learning approach can learn congestion control that maximizes this objective given a couple of minutes of time (without any pre-training) (see \autoref{fig:demo}).

\begin{figure}
\begin{minipage}{\columnwidth}
\subfloat[One sender]{\includegraphics[width=\columnwidth]{{{/home/max/repos/remy/compute-0-4.local/figures/4_1_2018-01-25-20:17:15_3_2018-01-25-20:17:15.0}}}}{}
\subfloat[Two senders]{\includegraphics[width=\columnwidth]{{{/home/max/repos/remy/compute-0-6.local/figures/4_2_2018-01-25-19:53:39_22_2018-01-25-19:53:39.0}}}}{}
\caption{Throughput and loss rate when sharing a bottleneck link of 20\,Mbit/s with a two-way end-to-end delay of 50\,ms and a very small buffer of $\frac{1}{10}$ bandwidth delay product between 1 and 2 senders (from top to bottom). The time needed for convergence increases with more senders, which is caused by the fact that the more senders use the link, the fewer packets each sender receives per unit of time for training and because of the increasing complexity of the environment. It is interesting to note that in the case of two senders, both flows complement each other and quickly adapt to the sending rate of the other flow.}
\label{fig:demo}
\end{minipage}
\end{figure}

After spending up to a few minutes with suboptimal performance during initial exploration (if no pre-training is used), our approach can compete with established algorithms and achieve superior throughput and packet loss of less that 1\%, while still gradually improving and adapting its way of performing congestion control.

@@

Congestion control has been implemented in TCP since the 1980s as a series of congestion collapses dramatically decreased applications' throughput in the Internet, which lead to the development of the Tahoe and Reno algorithms \cite{jacobson_congestion_1988}. These approaches as well as most others maintain a congestion window which stands for the maximum amount of data that are allowed to be traveling in the network without having been acknowledged by the receiver. In general, receiving acknowledgements of data that were successfully transmitted leads to the congestion window being increased while packet loss (or an increase in delay as in the Vegas algorithm \cite{brakmo_tcp_1995}) lead to the congestion window being lowered according to a set of fixed rules. Algorithms such as Cubic and Compound \cite{ha_cubic:_2008, tan_compound_2006} improve congestion control specifically with respect to so-called long fat networks with a high round-trip time and bandwidth. Proportional rate reduction \cite{dukkipati_proportional_2011} aims to make the reduction of the rate that occurs in case of congestion more smooth and steady in time to avoid the bursty behavior that occurred upon loss in previous TCP congestion control variants. On the other hand, BBR \cite{cardwell_bbr:_2016} does not use a set of fixed rules such as previous congestion control algorithms but instead estimates a model of the network path by using measurements of RTT and throughput and then adjusts its sending rate so that it uses the maximum bandwidth according to its network model while trying not to fill up queues. A similar approach is PCC \cite{dong_pcc:_2015}, which also uses measurements to find an optimum sending rate with respect to a defined utility function (e.g. maximizing throughput while minimizing packet loss). However, PCC doesn't maintain a congestion window but instead modulates the sending rate, using one constant sending rate for a certain time span.

Besides the aforementioned human-designed algorithms there have been attempts to use machine learning to improve congestion control. \citet{geurts_machine_2004} train a classifier off-line to determine whether packet loss is caused by congestion or by the link layer in case of wireless links. If it is caused by the link layer, TCP does not alter the window. Otherwise -- if a packet was lost due to congestion according to the classifier -- it uses a traditional TCP congestion control algorithm. \citet{shaio_reinforcement_2005} use online reinforcement learning to choose a sending rate for multimedia streaming. \citet{winstein_tcp_2013} train a machine learning solution called \textit{Remy} that finds an optimum congestion control for a given range of network parameters. For instance, one could find an optimum congestion control for networks with an RTT of 50-100\,ms and link speeds of 100-500\,Mbit/s. After a lengthy training procedure of several CPU weeks, Remy finds an optimum congestion control algorithm for the specified networks on a per acknowledgement basis (updating the window upon receiving an acknowledgement alike traditional congestion control algorithms).

Our goal is to find an optimum congestion control algorithm on a per acknowledgement basis similar to Remy that, however, can be trained on-line.  Such a solution could be used in a purely on-line fashion, in a off-line fashion like Remy or a combination of both: One could pre-train a generic congestion control algorithm that works reasonably well for every network that then gets refined during on-line training according to the current network circumstances.

To this end we propose Partial Action Learning which is based on the Asynchronous Advantage Actor Critic framework \cite{mnih_asynchronous_2016}, which has been demonstrated to be able to learn to play a wide range of video games and commonly outperform human players. In particular, it is a good choice for congestion control as it has been proven to deliver good performance and high training speed for playing video games and also for networking related tasks such as Adaptive Bitrate Algorithms \cite{mao_neural_2017} as its use of neural network make it able to even learn complex tasks.

@@

Our proposed framework \textit{Partial Action Learning} (PAL) (see \ref{subsec:pal}) framework is based on the Actor Critic framework for neural networks proposed by \citet{mnih_asynchronous_2016}, which we outline (with a focus on intuition over rigor) in this section.

There are two neural networks, the Actor Network and the Critic Network. Given a state, the Actor Network outputs what it deems to be the optimum action to perform in that certain state. The Critic Network estimates what long-term reward can be expected in this state. So an action is considered good if it achieved an empirically experienced long-term reward that is higher than the predicted long-term reward expected by the Critic Network and it is considered bad if the reward was lower than expected. The long-term reward is calculated as an exponentially weighted moving average of future rewards. So if a high reward can be achieved right now this is more favorable than if it can be achieved in the future. However it can also be beneficial to get a low reward now and instead get a very large one in the future.

@@

The Critic Network outputs the expected long-term reward $V(s_{t}; \theta)$ given a state $s_t$ at time step $t$ and the parameters (neural network weights) of the critic network $\theta$.

With $r_t$ being the reward that was received at time $t$ and $\gamma$ being a factor with $0 < \gamma \leq 1$, which stands for the influence that future rewards have on the moving average, we define the expected long-term reward at time step $t$ as 
\begin{align}
R_t = \gamma r_{t} + (1-\gamma) R_{t+1} ,
%R_t = \left(\left(\sum_{i=0}^{k-1} \gamma^ir_{t+i}\right) + \gamma^k V(s_{t+k}; \theta)\right)\left( 1-\gamma \right),
\end{align}

One can see that to compute $R_t$ at time step $t$ one has to look infinitely far into the future to get all future rewards as in each step one needs $R_{t+1}$. Thus, one usually only carries out $t_\text{max}$ steps (for example 20), waits for all the rewards and then calculates $R_t$ for each step and uses an estimation of $R_{t_\text{max}+1}$ provided by the critic network as the continuation in the future. The only exception from this is when the task to carry out is over before reaching $t_\text{max}$ (e.g. the TCP flow is over). In this case, one can either use a prediction from the critic network as a continuation as usually or simply let the moving average end without continuation i.e.~ignoring the term weighted with $(\gamma-1)$ at the last step.

@@

The loss function\footnote{not in the sense of packet loss but in the sense of the loss of a machine learning problem} (as defined by \cite{mnih_asynchronous_2016}), which the critic network tries to minimize at each time step, is the square of the difference of the actual long-term reward received and the expected long-term reward
\begin{align}
l_{\text{v},t} = \left(R_t - V(s_t; \theta)\right)^2.
\end{align}

@@

The Actor Network outputs a normal distribution parametrized by a mean $\mu$ and a standard deviation $\sigma$ from which the action $a_t$ at time step $t$ is randomly sampled. The main idea is that the neural network learns to output the right mean at the right time step to maximize the future reward and that it uses the standard deviation to try out new actions, which could yield a better than expected reward.

Each time we take an action $a_t$ we also let the critic network make an estimation $V(s_t; \theta)$, given the current neural network weights $\theta$ and the current state $s_t$. The loss function of the actor network can now be described as this: How likely was it to take action $a_t$ given the probability distribution ($\pi$) at time $t$ and how beneficial was this action when comparing it to the critic's estimation? For example, if an action was very successful and highly unlikely, the actor network's weights are updated to make actions like this more likely in the future.

The actor also aims at increasing the probability density function's entropy ($\frac{1}{2} \log \left( 2 \pi e \sigma^2 \right)$) to encourage exploration: Otherwise the actor could always perform similar actions but never actually take the action that yields the best performance. The actor network will aim to maximize the entropy and at the same time output the best possible action: If it realizes that the entropy is too large and bad actions are taken, it learns to lower the entropy. We designate the entropy as $H$ and introduce a parameter $\beta$ which specifies magnitude of the influence of the entropy. The loss function of the actor network is defined as follows \cite{mnih_asynchronous_2016}:
\begin{align}
\begin{split}
l_{\text{a},t} =& -\log \left( \pi\left( a_t \given s_t; \theta \right)\right)\left( R_t - v_t \right)\\ 
&- \beta H\left( \pi\left( s_t; \theta \right)\right).
\end{split}
\end{align}

@@

The key difference between PAL and previous approaches to RL is that in classical RL an action is always followed by a reward and a reward is always followed by an action. In our proposed concept, however, it is possible to take new actions while previous actions haven't been rewarded yet.

@@

In \autoref{alg:pal} (in the appendix) we show the code that runs in each of the agents (in the congestion control scenario, one agent corresponds to a sender). It is possible to have several agents which share a set of neural weights (which is usually done when performing off-line training \cite{mnih_asynchronous_2016,mao_neural_2017}) but one can also use separate weights for each agent, which is more realistic in case of congestion control in the Internet, as it is not sensible that different senders share a set of neural network weights over the Internet.

@@

\caption{A petri net describing the congestion control mechanism. We start with one token in the state \textit{Received feedback}, which means that we can take an action right in the beginning. The concept is the following: If there is at least one token in the \textit{Open congestion window} send a packet (for the sake of simplicity in this example we assume that the congestion window is in packets and not bytes). The \textit{Open congestion window} signifies the congestion window minus the packets that are currently unacknowledged. The token now goes to the network and after some time the acknowledgement (ACK) (or negative-acknowledgement (NACK) or timeout if the packet gets lost) for the packet is received in the \textit{Received ACK} state. From here one token goes to the \textit{Open congestion window} meaning that when we get an ACK, NACK (or timeout) we can send another packet. Furthermore the state \textit{Received feedback} receives a token upon receiving an ACK/NACK/timeout. \textit{Take action} adds \textit{a} tokens to the \textit{Open congestion window}, where $a$ is a real number (possibly also negative); thus in each state there can also be a real number of tokens (e.g.~ 2.34 tokens are possible in the \textit{Open congestion window}). However, we define that the congestion window can never be smaller than 1, which means that the sum of all tokens in the \textit{Open congestion window.}, \textit{Received ACK} and \textit{Packets in the network} states can never be smaller than 1.}
\label{fig:petri}
\end{figure}

@@

To use PAL for congestion control we first have to define the correct semantics for this specific use case and define what makes up the state and the reward. Furthermore, we have to define how the actor and critic network explicitly work in case of congestion control. In the following a time step $t$ corresponds to the reception of an acknowledgement. The beginning of the flow, before any packet is sent, corresponds to time step $0$.

\begin{description}
\item[$\textit{s}_\text{t}$] The state  describes the current ``congestion state'' defined by:
\begin{itemize}
\item the time between the last two packets that were received
\item the RTT of the last received packet
\item whether the last packet was lost, where $1$ denotes that the last packet was lost while $0$ indicates that it was received correctly
\item the current congestion window (as a real number, as actions by output by the actor network can be very small and thus the congestion window is often only changed by minuscule amounts)%
\end{itemize}
Of each of these features a exponentially weighted moving average (ewma) with a factor $\alpha$ of $\frac{1}{8}$ and one with an $\alpha$ of $\frac{1}{256}$ (the concept of using ewmas with these specific values of $\alpha$ was first introduced in \textit{Remy} \cite{winstein_tcp_2013}), which makes a total of 8 features.  
Each time an acknowledgement is received, the state is updated and the actor network is asked for the next action.
\item[$\textit{a}_t$] Based on a given state and the history of previous states (because we use Gated Recurrent Units (GRUs) \cite{cho_learning_2014} and thus can also consider previous states), the actor network returns an action $a_t$, which is a real number that stands for the change of the congestion window.

It might seem simplistic to only use an additive increase to the congestion window and no more sophisticated mechanisms such as a combination of a multiplicative increase and an additive increase. However, as we use GRUs in the actor network, which have a nonlinear output function (a hyperbolic tangent function), they can learn to model nonlinear relationships between the input and the output.  
\item[$\textit{r}_t$] The reward is a tuple of three reward metrics (defined in the next paragraph). 
%These are discussed in more detail in \autoref{subsubsec:critic}. 
For each reward metric there is also an output of the critic network that predicts the expected long-term average of this reward metric given the current state. 
\item[$\textit{v}_t$] The value is a tuple of the expected average reward estimated by the critic network (see \autoref{subsubsec:critic}) for each of the reward metrics. 
\end{description}

>>>>

%Conference
%\acmConference[WOODSTOCK'97]{ACM Woodstock conference}{July 1997}{El
%  Paso, Texas USA} 
%\acmYear{2018}
%\copyrightyear{2018}

@@

\begin{document}
\title{TCP Rax: Deep Reinforcement Learning for Congestion Control}
%\titlenote{Produces the permission block, and
%  copyright information}
%\subtitle{Extended Abs\texttt{tract}
%\subtitlenote{The full version of the author's guide is available as
%  \texttt{acmart.pdf} document}}

@@

\renewcommand{\shortauthors}{X. et al.}

@@

This paper proposes Reactive Adaptive eXperience based congestion control (TCP Rax), a new method of congestion control that uses online reinforcement learning to maintain an optimum congestion window with respect to a given utility function and based on current network conditions. We use a neural network based approach that can be initialized either with random weights (without prior knowledge) or with a previously trained neural network to improve stability and convergence time. The processing of rewards in congestion control depends on the arrival of acknowledgments, inducing loop delays that may lead to oscillations. As a remedy we propose \textit{Partial Action Learning} (PAL), a specific formulation of neural network based reinforcement learning that supports the processing of delayed and partial rewards.

We show that our method converges to a stable, close-to-optimum solution within the order of minutes and outperforms existing congestion control algorithms in typical networks. Thus, this paper demonstrates that neural network based reinforcement learning without any prior knowledge can feasibly be done online and can compete with human-designed congestion-control schemes such as New Reno and Cubic.

@@

In general, in congestion control each sender aims to maximize an objective function that usually comprises one or more of the following subgoals:
\begin{enumerate*}
\item Send as many packets as possible
\item with as little delay as possible
\item while minimizing the amount of lost packets
\end{enumerate*}. 
As, for instance, for some applications overall throughput is more important than little packet loss, each application can define its preference using a custom utility function. Besides aiming to fulfill these subgoals, it is also desirable from an overall point of view that all senders use available resources in a way that is fair to others: When several senders share an Internet link under the same conditions, they should receive an equal share of the available resources. To achieve fairness, it is necessary that senders only use utility functions that 
\begin{enumerate*}
\item do not give an unfair advantage over others
\item and converge to a stable equilibrium
\end{enumerate*}.

In a machine learning based congestion control method, each sender uses a set of observed environment conditions to determine which action it should take to maximize his reward in the future. The environment conditions are metrics that the sender can obtain from the network, for example the mean round-trip time of the last received packets, the packet loss rate etc.~or combinations thereof. An action is a change to the sending rate: For example, if a sender perceives an increase in the loss rate it might decide to lower the sending rate. To modulate the sending rate one can  either
\begin{enumerate*}
\item change the sending rate itself at certain points in time or
\item maintain a congestion window that indicates the number of packets that can be in flight at a certain point in time
\end{enumerate*}. As the operating system (OS) has to schedule different applications and share the network interfaces between them, maintaining a congestion window is appealing, because one cannot be sure that the OS can maintain a specific constant sending rate for each application on one machine.

While congestion control seems to be a well-suited task for reinforcement learning, there are a couple of unique aspects and differences that distinguish it from most other reinforcement learning problems:  \begin{itemize}
\item \textbf{Delayed rewards}: In congestion control, rewards are always delayed by one round-trip time. In most other areas where reinforcement learning is applied, if an action is performed, the consequences of that action become apparent immediately. Also, due to the delay, by the time an action receives its reward, probably other actions have already been performed. Thus, contrary to most RL applications, actions and rewards are not synchronized. 
\item \textbf{Actions and rewards are not atomic}: If an action causes three packets to be sent, we consider these three packets being sent three \textit{partial actions}. In case these packets are transmitted correctly, the receiver will send back three acknowledgements. Each of these acknowledgements is a \textit{partial reward} that unveils new information on the current state of the network and enables the sender to take an action based on this new information. However, only if all partial rewards have been received, the sender can assemble the overall reward of the corresponding action and can perform an update of the underlying reinforcement learning logic. 
\item \textbf{Variable number of actions and rewards}: If the congestion window is larger, more packets are sent and more acknowledgements are received, which means that the reinforcement learning algorithm gets more training data the larger the window is. This is problematic as it means that -- without proper adjustments -- the experience made at a larger congestion window will ``overwrite'' the experience made at a smaller window. With the larger window more training data arrives and it is thus going to be overemphasized. 
\end{itemize}

@@

While using PAL to train an optimum congestion control for a specific range of network scenarios in an offline fashion is possible, offline learning of an optimum TCP congestion control is something that has already been done previously in an approach called \textit{Remy} \citep{winstein_tcp_2013}. The only potential advantage that PAL can provide over prior solutions is increased training speed. Thus, we want to show that it is  not only possible to learn congestion control by using machine learning but that it is even possible to do so online and without any preknowledge about the network environment.

\autoref{fig:demo} shows that using a utility function that encourages high throughput and discourages packet loss (adapted from \citep{dong_pcc:_2015}) our machine learning approach can learn congestion control that maximizes this objective given a couple of minutes of time (without any pre-training).

\begin{figure}
\begin{minipage}{\columnwidth}
\subfloat[One sender\label{fig:demo1}]{\includegraphics[width=\columnwidth]{{{/home/max/repos/remy/compute-0-4.local/figures/4_1_2018-01-25-20:17:15_3_2018-01-25-20:17:15.0}}}}{}
\subfloat[Two senders\label{fig:demo2}]{\includegraphics[width=\columnwidth]{{{/home/max/repos/remy/compute-0-6.local/figures/4_2_2018-01-25-19:53:39_22_2018-01-25-19:53:39.0}}}}{}
\caption{Throughput and loss rate when sharing a bottleneck link of 20\,Mbit/s with a two-way end-to-end delay of 50\,ms and a very small buffer of $\frac{1}{10}$ bandwidth delay product between one (\ref{fig:demo1}) and two (\ref{fig:demo2}) senders . The time needed for convergence increases with more senders, which is caused by the fact that the more senders use the link, the fewer packets each sender receives per unit of time for training and because of the increasing complexity of the environment. In case of two senders, both flows complement each other and quickly adapt to the sending rate of the other flow.}
\label{fig:demo}
\end{minipage}
\end{figure}

After spending up to a few minutes with suboptimal performance during initial exploration (if no pre-training is used), our approach can compete with established algorithms and achieve superior throughput and an average packet loss of less than 1\%, while still gradually improving and adapting its way of performing congestion control.

Thus we make the three contributions:
\begin{itemize}
\item A window based TCP congestion control that acts on a per-acknowledgement basis that can compete with existing congestion control algorithms.
\item The ability to learn online, offline or a combination of both offline pre-training with online learning. 
\item A framework for deep reinforcement learning that allows dealing with delayed and partial rewards.
\end{itemize}

@@

Congestion control has been implemented in TCP since the 1980s as a series of congestion collapses dramatically decreased applications' throughput in the Internet, which lead to the development of the Tahoe and Reno algorithms \cite{jacobson_congestion_1988}. These approaches as well as most others maintain a congestion window which stands for the maximum amount of data that are allowed to be in flight in the network without having been acknowledged by the receiver. In most TCP congestion control algorithms, receiving acknowledgements of data that were successfully transmitted leads to the congestion window being increased while packet loss (or an increase in delay as in the Vegas algorithm \cite{brakmo_tcp_1995}) lead to the congestion window being lowered according to a set of fixed rules. Algorithms such as Cubic and Compound \cite{ha_cubic:_2008, tan_compound_2006} improve congestion control specifically with respect to so-called long fat networks with a high round-trip time and bandwidth. Proportional rate reduction \cite{dukkipati_proportional_2011} aims to make the reduction of the rate that occurs in case of congestion more smooth and steady in time to avoid the bursty behavior that occurred upon loss in previous TCP congestion control variants. A different approach is taken by BBR \cite{cardwell_bbr:_2016}, which does not use a set of fixed rules such as previous congestion control algorithms. Instead it estimates a model of the network path by using measurements of RTT and throughput and then adjusts its sending rate so that it uses the maximum bandwidth according to its network model while trying not to fill up queues. A similar approach is PCC \cite{dong_pcc:_2015}, which also uses measurements to find an optimum sending rate with respect to a defined utility function (e.g. maximizing throughput while minimizing packet loss). However, PCC doesn't maintain a congestion window but instead modulates the sending rate, using one constant sending rate for a certain time span.

Besides the aforementioned human-designed algorithms there have been attempts to use machine learning to improve congestion control. \citet{geurts_machine_2004} train a classifier offline to determine whether packet loss is caused by congestion or by the link layer in case of wireless links. If it is caused by the link layer, TCP does not alter the window. Otherwise -- if a packet was lost due to congestion according to the classifier -- it uses a traditional TCP congestion control algorithm. \citet{shaio_reinforcement_2005} use online reinforcement learning to choose a sending rate for multimedia streaming. \citet{winstein_tcp_2013} train a machine learning solution called \textit{Remy} that finds an optimum congestion control for a given range of network parameters. For instance, one could find an optimum congestion control for networks with an RTT of 50-100\,ms and link speeds of 10-20\,Mbit/s. After a lengthy training procedure of several CPU weeks, Remy finds an optimum congestion control algorithm for the specified networks on a per acknowledgement basis (updating the window upon receiving an acknowledgement like traditional congestion control algorithms).

Our goal is to find an optimum congestion control algorithm on a per acknowledgement basis similar to Remy that, however, can be trained online.  Such a solution could be used in a purely online fashion, in a offline fashion like Remy or a combination of both: One could pre-train a generic congestion control algorithm that works reasonably well for every network that then gets refined during online training according to the current network circumstances.

To this end we propose Partial Action Learning which is based on the Asynchronous Advantage Actor Critic framework \cite{mnih_asynchronous_2016}, which has been demonstrated to be able to learn to play a wide range of video games and commonly outperform human players. Its use of a neural network enable it to even learn complex tasks. In particular, it is a good choice for congestion control as it has been proven to deliver good performance and high training speed for playing video games and also for networking related tasks such as Adaptive Bitrate Algorithms \cite{mao_neural_2017}.

@@

Our proposed framework \textit{Partial Action Learning} (PAL) (see \ref{subsec:pal}) is based on the Actor Critic framework for neural networks proposed by \citet{mnih_asynchronous_2016}, which we outline (with a focus on intuition over rigor) in this section.

There are two neural networks, the Actor Network and the Critic Network. Given a state, the Actor Network outputs what it deems to be the optimum action to perform in that certain state. The Critic Network predicts what long-term reward can be expected in this state. So an action is considered good if it achieved an empirically experienced long-term reward that is higher than the predicted long-term reward expected by the Critic Network and it is considered bad if the reward was lower than expected. The long-term reward is calculated as an exponentially weighted moving average of future rewards. So if a high reward can be achieved right now this is more favorable than if it can be achieved in the future. However it can also be beneficial to get a low reward now and instead get a very large one in the future.

@@

The Critic Network outputs a prediction of the long-term reward $v_t = V(s_{t}; \theta)$ given a state $s_t$ at time step $t$ and the parameters (neural network weights) of the critic network $\theta$.

With $r_t$ being the reward that was received at time $t$ and $\gamma$ being a factor with $0 < \gamma \leq 1$, which stands for the influence that future rewards have on the moving average, we define the actual experienced long-term reward $R_t$ at time step $t$ as 
\begin{align}
R_t = \gamma r_{t} + (1-\gamma) R_{t+1} ,
%R_t = \left(\left(\sum_{i=0}^{k-1} \gamma^ir_{t+i}\right) + \gamma^k V(s_{t+k}; \theta)\right)\left( 1-\gamma \right).
\end{align}

To compute $R_t$ at time step $t$ it is necessary to look infinitely far into the future to get all future rewards as in each step $t$ to compute $R_t$ one needs $R_{t+1}$. Thus, one usually only carries out $t_\text{max}$ steps (for example 20), waits for all the rewards and then calculates $R_t$ for each step and uses an estimate of $R_{t_\text{max}+1}$ provided by the critic network as the continuation in the future. The only exception from this is when the task to execute ends before reaching $t_\text{max}$ (e.g. the TCP flow is over). In this case, one can either use a prediction from the critic network as a continuation or alternatively let the moving average end without continuation i.e.~ignoring the term weighted with $(\gamma-1)$ at the last step.

@@

The loss function\footnote{not in the sense of packet loss but in the sense of the loss function of a machine learning problem, which one aims to minimize} (as defined by \cite{mnih_asynchronous_2016}), which the critic network tries to minimize at each time step, is the square of the difference of the actual experienced long-term reward received and the predicted long-term reward $l_{\text{v},t}$ as
\begin{align}
l_{\text{v},t} = \left(R_t - V(s_t; \theta)\right)^2.
\end{align}

@@

The Actor Network outputs a normal distribution of the possible values for the action parametrized by a mean $\mu$ and a standard deviation $\sigma$ from which the action $a_t$ at time step $t$ is randomly sampled. The main idea is that the neural network learns to output the right mean of the actions at the right time step to maximize the future reward and that it uses the standard deviation to try out new actions, which could yield a better than expected reward.

Each time we take an action $a_t$ we also let the critic network make an estimation $v_t = V(s_t; \theta)$, given the current neural network weights $\theta$ and the current state $s_t$. The loss function of the actor network can be described as follows: How likely was it to take action $a_t$ given the probability distribution ($\pi$) at time $t$ and how beneficial was this action when comparing it to the critic's estimation? For example, if an action was very successful and highly unlikely, the actor network's weights are updated to make actions like this more likely in the future.

The actor also aims at increasing the probability density function's entropy ($H\left( \pi\left( s_t; \theta \right)\right) = \frac{1}{2} \log \left( 2 \pi e \sigma^2 \right)$) to encourage exploration: Otherwise the actor could always perform similar actions but never actually take the action that yields the best performance. The actor network will aim to maximize the entropy and at the same time output the best possible action: If it realizes that the entropy is too large and bad actions are taken, it learns to lower the entropy. We designate the entropy as $H$ and introduce a parameter $\beta$ which specifies the magnitude of the influence of the entropy. The loss function of the actor network is defined as follows \cite{mnih_asynchronous_2016}:
\begin{align}
\begin{split}
l_{\text{a},t} =& -\log \left( \pi\left( a_t \given s_t; \theta \right)\right)\left( R_t - V(s_t; \theta) \right)\\ 
&- \beta H\left( \pi\left( s_t; \theta \right)\right).
\end{split}
\end{align}

@@

The key difference between our approach PAL and previous approaches to RL is that in classical RL an action is always followed by a reward and a reward is always followed by an action. In our proposed concept, however, it is possible to take new actions while previous actions haven't been rewarded yet.

@@

In \autoref{alg:pal} (in the appendix) we show the code that runs in each of the agents (in the congestion control scenario, one agent corresponds to a sender). It is possible to have several agents which share a set of neural weights (which is usually done when performing offline training \cite{mnih_asynchronous_2016,mao_neural_2017}) but one can also use separate weights for each agent, which is more realistic in case of congestion control in the Internet, as it is not sensible that different senders share a set of neural network weights over the Internet.

@@

\caption{A petri net describing the congestion control mechanism. It starts with one token in the state \textit{Received feedback}, which means that it can take an action right in the beginning. The concept is the following: If there is at least one token in the \textit{Open congestion window} it sends a packet. For the sake of simplicity in this example we assume that the congestion window is in packets and not bytes. The \textit{Open congestion window} designates the congestion window minus the packets that are currently unacknowledged. The token now is passed to the network and after some time the acknowledgement (ACK) (or negative-acknowledgement (NACK) or timeout if the packet gets lost) for the packet is received in the \textit{Received ACK} state. From here one token goes to the \textit{Open congestion window} meaning that when it receives an ACK, NACK (or timeout) it can send another packet. Furthermore the state \textit{Received feedback} receives a token upon receiving an ACK/NACK/timeout. \textit{Take action} adds \textit{a} tokens to the \textit{Open congestion window}, where $a$ is a real number (possibly also negative). Therefore, in each state there can be a real number of tokens (e.g.~ 2.34 tokens are possible in the \textit{Open congestion window}). However, we define that the congestion window can never be smaller than 1, which means that the sum of all tokens in the \textit{Open congestion window.}, \textit{Received ACK} and \textit{Packets in the network} states can never be smaller than 1.}
\label{fig:petri}
\end{figure}

@@

To use reinforcement learning for congestion control in general and our modified approach PAL specifically we first have to define the correct semantics for this specific use case and define what makes up the state and the reward. Furthermore, we have to define how the actor and critic network explicitly work in case of congestion control. In the following, a time step $t$ corresponds to the reception of an acknowledgement. The beginning of the flow, the time when the first packet is sent, corresponds to time step $0$.

\begin{description}
\item[$\textit{s}_\text{t}$] The state $s_t$ describes the current ``congestion state'' defined by:
\begin{itemize}
\item the time between the last two acknowledgment that were received
\item the RTT of the last received packet
\item a loss indicator that shows whether the last packet was lost, where $1$ denotes that the last packet was lost while $0$ indicates that it was received correctly
\item the current congestion window (as a real number, as actions by output by the actor network can be very small and thus the congestion window is often changed by minuscule amounts)%
\end{itemize}
These features are not used themselves but instead of each of these features an exponentially weighted moving average (ewma) with a factor $\alpha$ of $\frac{1}{8}$ and one with an $\alpha$ of $\frac{1}{256}$ is computed, which makes a total of 8 features. The concept of using ewmas with these specific values of $\alpha$ was first introduced in \textit{Remy} \cite{winstein_tcp_2013}).
Each time an acknowledgement is received, the state is updated and the actor network is asked for the next action.
\item[$\textit{a}_t$] The action $a_t$ is a real number that represents the change to the congestion window. It is computed based on a given state and the history of previous states. The history of previous states as considered as our neural network uses Gated Recurrent Units (GRUs) \cite{cho_learning_2014}.

It might seem simplistic to only use an additive increase to the congestion window and no more sophisticated mechanisms such as a combination of a multiplicative increase and an additive increase. However, GRUs have a nonlinear output function (a hyperbolic tangent function) and thus they can learn to model nonlinear relationships between the input and the output. 
\item[$\textit{r}_t$] The reward $r_t$ is a tuple of three reward metrics (defined in the next paragraph). 
%These are discussed in more detail in \autoref{subsubsec:critic}. 
For each reward metric there is also an output of the critic network that predicts the expected long-term average of this reward metric given the current state. 
\item[$\textit{v}_t$] The value $v_t$ is a tuple of the expected average reward estimated by the critic network (see \autoref{subsubsec:critic}) for each of the reward metrics. Also for the value $v_t$ the current state and the history of previous states is considered. 
\end{description}
